---
title: "Census Datasets example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{census_datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 80)
```

<!--
THIS VIGNETTE IS BASED ON:
https://github.com/chanzuckerberg/cell-census/blob/main/api/python/notebooks/api_demo/census_datasets.ipynb
-->

*Goal:* demonstrate basic use of the `census_datasets` data frame.

Each Cell Census contains a top-level data frame itemizing the datasets contained therein. You can read this `SOMADataFrame` into an [Arrow Table](https://arrow.apache.org/docs/r/reference/Table.html): 

```{r}
census <- CellCensus::open_soma()
census_datasets <- census$get("census_info")$get("datasets")$read()
print(census_datasets)
```

and then an R data frame:

```{r}
census_datasets <- as.data.frame(census_datasets)
print(census_datasets[, c(
  "dataset_id",
  "dataset_title",
  "dataset_total_cell_count"
)], n = 5)
```

The sum of cell counts across all datasets should match the number of cells across all SOMA experiments (human, mouse).

```{r}
census_data <- census$get("census_data")
all_experiments <- lapply(census_data$to_list(), function(it) census_data$get(it$name))
print(all_experiments)
```

```{r}
experiments_total_cells <- sum(sapply(all_experiments, function(it) {
  nrow(it$obs$read(column_names = c("soma_joinid")))
}))
print(paste("Found", experiments_total_cells, "cells in all experiments."))
print(paste(
  "Found", sum(as.vector(census_datasets$dataset_total_cell_count)),
  "cells in all datasets."
))
```

Let's pick one dataset to slice out of the census, and turn into a Seurat in-memory object. This can be used with the Seurat toolchain immediately, or stored locally.

```{r}
census_datasets[census_datasets$dataset_id == "0bd1a1de-3aee-40e0-b2ec-86c7a30c7149", ]
```

Create a query on the mouse experiment, "RNA" measurement, for the `dataset_id`.

```{r}
obs_query <- tiledbsoma::SOMAAxisQuery$new(
  value_filter = "dataset_id == '0bd1a1de-3aee-40e0-b2ec-86c7a30c7149'"
)
expt_query <- tiledbsoma::SOMAExperimentAxisQuery$new(
  census_data$get("mus_musculus"), "RNA",
  obs_query = obs_query
)
print(expt_query)
# FIXME executing expt_query:
#   Currently unspported type: UTF8
#   Calls: <Anonymous> ... tiledb_query_condition_init -> libtiledb_query_condition_init
# TODO: to_seurat
```

TODO: get_h5ad_uri

