---
title: "Genes measured in each cell (dataset presence matrix)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genes measured in each cell (dataset presence matrix)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 88, max.print = 100)
```

<!--
THIS VIGNETTE IS BASED ON:
https://github.com/chanzuckerberg/cellxgene-census/blob/main/api/python/notebooks/api_demo/census_dataset_presence.ipynb
-->

The Census is a compilation of cells from multiple datasets that may differ by the sets of genes they measure. This notebook describes the way to identify the genes measured per dataset.

The presence matrix is a sparse boolean array, indicating which features (var) were present in each dataset. The array has dimensions [n_datasets, n_var], and is stored in the SOMA Measurement `varp` collection. The first dimension is indexed by the `soma_joinid` in the `census_datasets` dataframe. The second is indexed by the `soma_joinid` in the `var` dataframe of the measurement.

As a reminder the `obs` data frame has a column `dataset_id` that can be used to link any cell in the Census to the presence matrix.

Contents

1. Opening the Census.
2. Fetching the IDs of the Census datasets.
3. Fetching the dataset presence matrix.
4. Identifying genes measured in a specific dataset.
5. Identifying datasets that measured specific genes.
6. Identifying all genes measured in a dataset.

## Opening the Census

The `cellxgene.census` R package contains a convenient API to open any version of the Census (by default, the newest stable version).

```{r,message=FALSE,warning=FALSE}
library("cellxgene.census")
census <- open_soma()
```

## Fetching the IDs of the Census datasets

Let's grab a table of all the datasets included in the Census and use this table in combination with the presence matrix below.

```{r}
# Grab the experiment containing human data, and the measurement therein with RNA
human <- census$get("census_data")$get("homo_sapiens")
human_rna <- human$ms$get("RNA")

# The census-wide datasets
datasets_df <- as.data.frame(census$get("census_info")$get("datasets")$read()$concat())
print(datasets_df)
```

## Fetching the dataset presence matrix

Now let's fetch the dataset presence matrix.

For convenience, read the entire presence matrix (for Homo sapiens) into a sparse matrix. There is a convenience function providing this capability:

```{r}
presence_matrix <- get_presence_matrix(census, "Homo sapiens", "RNA")
print(dim(presence_matrix))
```

We also need the `var` dataframe, which is read into an R data frame for convenient manipulation:

```{r}
var_df <- as.data.frame(human_rna$var$read()$concat())
print(var_df)
```

## Identifying genes measured in a specific dataset

Now that we have the dataset table, the genes metadata table, and the dataset presence matrix, we can check if a gene or set of genes were measured in a specific dataset.

**Important:** the presence matrix is indexed by `soma_joinid`, and is *NOT* positionally indexed. In other words:

- the first dimension of the presence matrix is the dataset's `soma_joinid`, as stored in the `census_datasets` dataframe.
- the second dimension of the presence matrix is the feature's `soma_joinid`, as stored in the `var` dataframe.

The presence matrix has a method `$take()` that lets you slice it by `soma_joinid`s from `census_datasets` and `var`. And the full presence matrix, or slices of it, can then be exported to a regular matrix with the method `$get_one_based_matrix()`

Let's find out if the the gene `"ENSG00000286096"` was measured in the dataset with id `"97a17473-e2b1-4f31-a544-44a60773e2dd"`.

```{r}
# Get soma_joinid for datasets and genes of interest
var_joinid <- var_df$soma_joinid[var_df$feature_id == "ENSG00000286096"]
dataset_joinid <- datasets_df$soma_joinid[datasets_df$dataset_id == "97a17473-e2b1-4f31-a544-44a60773e2dd"]

# Slice presence matrix with datasets and genes of interest
presence_matrix_slice <- presence_matrix$take(i = dataset_joinid, j = var_joinid)

# Convert presence matrix to regular matrix
presence_matrix_slice <- presence_matrix_slice$get_one_based_matrix()

# Find how if the gene is present in this dataset
is_present <- presence_matrix_slice[, , drop = TRUE]
cat(paste("Feature is", if (is_present) "present." else "not present."))
```

## Identifying datasets that measured specific genes

Similarly, we can determine the datasets that measured a specific gene or set of genes.

```{r}
# Grab the feature's soma_joinid from the var dataframe
var_joinid <- var_df$soma_joinid[var_df$feature_id == "ENSG00000286096"]

# The presence matrix is indexed by the joinids of the dataset and var dataframes,
# so slice out the feature of interest by its joinid.
presence_matrix_slice  <- presence_matrix$take(j = var_joinid)$get_one_based_matrix()
measured_datasets <- presence_matrix_slice[, , drop = TRUE] != 0
dataset_joinids <- datasets_df$soma_joinid[measured_datasets]

# From the datasets dataframe, slice out the datasets which have a joinid in the list
print(datasets_df[dataset_joinids, ])
```

## Identifying all genes measured in a dataset

Finally, we can find the set of genes that were measured in the cells of a given dataset.

```{r}
# Slice the dataset(s) of interest, and get the joinid(s)
dataset_joinids <- datasets_df$soma_joinid[datasets_df$collection_id == "17481d16-ee44-49e5-bcf0-28c0780d8c4a"]

# Slice the presence matrix by the first dimension, i.e., by dataset
presence_matrix_slice <- presence_matrix$take(i = dataset_joinids)$get_one_based_matrix()
genes_measured <- Matrix::colSums(presence_matrix_slice) > 0
var_joinids <- var_df$soma_joinid[genes_measured]

print(var_df[var_joinids, ])
```

### Close the census

After use, the census object should be closed to release memory and other resources.

```{r}
census$close()
```

This also closes all SOMA objects accessed via the top-level `census`. Closing can be automated using `on.exit(census$close(), add = TRUE)` immediately after `census <- open_soma()`.
