---
title: "Generating citations for Census slices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating citations for Census slices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 88, max.print = 256)
```

This notebook demonstrates how to generate a citation string for all datasets contained in a census slice.

**Contents**

1. Requirements
2. Generating citation strings
   1. Via cell metadata query
   2. Via Seurat query
   3. Via SingleCellExperiment query

⚠️ Note that the Census RNA data includes duplicate cells present across multiple datasets. Duplicate cells can be filtered in or out using the cell metadata variable `is_primary_data` which is described in the [Census schema](https://github.com/chanzuckerberg/cellxgene-census/blob/main/docs/cellxgene_census_schema.md#repeated-data).

## Requirements

This notebook requires:

- `cellxgene_census` Python package.
- Census data release with [schema version](https://github.com/chanzuckerberg/cellxgene-census/blob/main/docs/cellxgene_census_schema.md) 1.3.0 or greater.

## Generating citation strings

First we open a handle to the Census data. To ensure we open a data release with schema version 1.3.0 or greater, we use `census_version="latest"`

```{r,message=FALSE,warning=FALSE}
library("tiledb")
library("cellxgene.census")

census <- open_soma(census_version = "latest")
census_release_info <- census$get("census_info")$get("summary")$read()$concat()
as.data.frame(census_release_info)
```

Then we load the dataset table which contains a column `"citation"` for each dataset included in Census. 

```{r,message=FALSE,warning=FALSE}
datasets <- census$get("census_info")$get("datasets")$read()$concat()
datasets <- as.data.frame(datasets)
head(datasets["citation"])
```

And now we can use the column `"dataset_id"` present in both the dataset table and the Census cell metadata to create citation strings for any Census slice.

### Via cell metadata query

```{r,message=FALSE,warning=FALSE}
# Query cell metadata
cell_metadata <- census$get("census_data")$get("homo_sapiens")$obs$read(
    value_filter = "tissue == 'cardiac atrium'",
    column_names = c("dataset_id", "cell_type")
)

cell_metadata <- as.data.frame(cell_metadata$concat())

# Get a citation string for the slice
slice_datasets = datasets[datasets$dataset_id  %in% cell_metadata$dataset_id,]
print(slice_datasets$citation)
```

### Via Seurat query

```{r,message=FALSE,warning=FALSE}
# Fetch a Seurat object
seurat_obj <- get_seurat(
    census = census,
    organism = "homo_sapiens",
    measurement_name = "RNA",
    obs_value_filter = "tissue == 'cardiac atrium'",
    var_value_filter = "feature_name == 'MYBPC3'",
    obs_column_names = c("dataset_id", "cell_type")
)

# Get a citation string for the slice
slice_datasets = datasets[datasets$dataset_id  %in% seurat_obj[[]]$dataset_id,]
print(slice_datasets$citation)
```

### Via SingleCellExperiment query

```{r,message=FALSE,warning=FALSE}
# Fetch a Seurat object
sce_obj <- get_single_cell_experiment(
    census = census,
    organism = "homo_sapiens",
    measurement_name = "RNA",
    obs_value_filter = "tissue == 'cardiac atrium'",
    var_value_filter = "feature_name == 'MYBPC3'",
    obs_column_names = c("dataset_id", "cell_type")
)

# Get a citation string for the slice
slice_datasets = datasets[datasets$dataset_id  %in% sce_obj$dataset_id,]
print(slice_datasets$citation)
```
