---
title: "Census datasets presence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Census datasets presence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 80)
```

<!--
THIS VIGNETTE IS BASED ON:
https://github.com/chanzuckerberg/cellxgene-census/blob/main/api/python/notebooks/api_demo/census_dataset_presence.ipynb
-->

*Goal:* demonstrate basic use of the `datasets_presence_matrix` array.

The presence matrix is a sparse array, indicating which features (var) were present in each dataset. The array has dimensions [n_datasets, n_var], and is stored in the SOMA Measurement `varp` collection. The first dimension is indexed by the `soma_joinid` in the `census_datasets` dataframe. The second is indexed by the `soma_joinid` in the `var` dataframe of the measurement.

```{r}
census <- cellxgene.census::open_soma()
# Grab the experiment containing human data, and the measurement therein with RNA
human <- census$get("census_data")$get("homo_sapiens")
human_rna <- human$ms$get("RNA")

# The census-wide datasets
datasets_df <- as.data.frame(census$get("census_info")$get("datasets")$read())
print(datasets_df)
```

For convenience, read the entire presence matrix (for Homo sapiens) into a `Matrix::sparseMatrix`. There is a convenience API providing this capability:

```{r}
presence_matrix <- cellxgene.census::get_presence_matrix(census, "Homo sapiens", "RNA")
print(dim(presence_matrix))
```

We also need the `var` dataframe, which is read into an R data frame for convenient manipulation:

```{r}
var_df <- as.data.frame(human_rna$var$read())
print(var_df)
```

## Is a feature present in a dataset?

*Goal:* test if a given feature is present in a given dataset.

**Important:** the (one-based) indexes in the sparse presence matrix correspond to the (zero-based) `soma_joinid` + 1. In other words:

- the first dimension of the presence matrix is (one plus) the dataset's `soma_joinid` as stored in the `census_datasets` dataframe.
- the second dimension of the presence matrix is (one plus) the feature's `soma_joinid` as stored in the `var` dataframe.

```{r}
var_joinid <- var_df$soma_joinid[var_df$feature_id == "ENSG00000286096"]
dataset_joinid <- datasets_df$soma_joinid[datasets_df$dataset_id == "97a17473-e2b1-4f31-a544-44a60773e2dd"]
is_present <- presence_matrix[dataset_joinid + 1, var_joinid + 1]
cat(paste("Feature is", if (is_present) "present." else "not present."))
```

## What datasets contain a feature?

*Goal:* look up all datasets that have a feature_id present.

```{r}
# Grab the feature's soma_joinid from the var dataframe
var_joinid <- var_df$soma_joinid[var_df$feature_id == "ENSG00000286096"]

# The presence matrix is indexed by the joinids of the dataset and var dataframes,
# so slice out the feature of interest by its joinid.
dataset_joinids <- datasets_df$soma_joinid[presence_matrix[, var_joinid + 1] != 0]

print(datasets_df[dataset_joinids + 1, ])
```

## What features are in a dataset?


*Goal:* lookup the features present in a given dataset.

This example also demonstrates the ability to do the query on multiple datasets.

```{r}
# Slice the dataset(s) of interest, and get the joinid(s)
dataset_joinids <- datasets_df$soma_joinid[datasets_df$collection_id == "17481d16-ee44-49e5-bcf0-28c0780d8c4a"]

# Slice the presence matrix by the first dimension, i.e., by dataset
var_joinids <- var_df$soma_joinid[which(Matrix::colSums(presence_matrix[dataset_joinids + 1, ]) > 0)]

print(var_df[var_joinids + 1, ])
```
