---
title: "Axis Query Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{census_axis_query}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 80)
```

<!--
THIS VIGNETTE IS BASED ON:
https://github.com/chanzuckerberg/cell-census/blob/main/api/python/notebooks/api_demo/census_axis_query.ipynb
-->

# Axis Query Example

*Goal:* demonstrate basic axis metadata handling.

The CZ CELLxGENE Census stores obs (cell) metadata in a SOMA DataFrame, which can be queried and read as an R data frame. The Cell Census also has a convenience package which simplifies opening the census.

R data frames are in-memory objects. Take care that queries are small enough for results to fit in memory.

## Opening the census

The `cellxgene.census` R package contains a convenient API to open the latest version of the Census.

```{r}
census <- cellxgene.census::open_soma()
```

You can learn more about the cellxgene.census methods by accessing their corresponding documentation. For example `?cellxgene.census::open_soma`.

## Summarize Census cell metadata

Tips:

* You can read an *entire* SOMA dataframe into R using `as.data.frame(soma_df$read())`.
* Queries will be much faster if you request only the DataFrame columns required for your analysis (e.g. `column_names = c("soma_joinid", "cell_type_ontology_term_id")`).
* You can also further refine query results by using a `value_filter`, which will filter the census for matching records.

### Summarize all cell types

This example reads the cell metadata (obs) into an R data frame to summarize in a variety of ways.

```{r}
human <- census$get("census_data")$get("homo_sapiens")

# Read obs into an R data frame (tibble).
obs_df <- as.data.frame(human$obs$read(
  column_names = c("soma_joinid", "cell_type_ontology_term_id")
))

# Find all unique values in the cell_type_ontology_term_id column.
unique_cell_type_ontology_term_id <- unique(obs_df$cell_type_ontology_term_id)

cat(paste(
  "There are",
  length(unique_cell_type_ontology_term_id),
  "cell types in the Cell Census! The first few are:"
))
head(unique_cell_type_ontology_term_id)
```

### Summarize a subset of cell types, selected with a `value_filter`

This example utilizes a SOMA "value filter" to read the subset of cells with `tissue_ontology_term_id` equal to `UBERON:0002048` (lung tissue), and summarizes the query result.

```{r}
# Read cell_type terms for cells which have a specific tissue term
LUNG_TISSUE <- "UBERON:0002048"

obs_df <- as.data.frame(human$obs$read(
  column_names = c("cell_type_ontology_term_id"),
  value_filter = paste("tissue_ontology_term_id == '", LUNG_TISSUE, "'", sep = "")
))

# Find all unique values in the cell_type_ontology_term_id column as an R data frame.
unique_cell_type_ontology_term_id <- unique(obs_df$cell_type_ontology_term_id)
cat(paste(
  "There are ",
  length(unique_cell_type_ontology_term_id),
  " cell types in the Cell Census where tissue_ontology_term_id == ",
  LUNG_TISSUE,
  "!\nThe first few are:",
  sep = ""
))
head(unique_cell_type_ontology_term_id)

# Report the 10 most common
top_10 <- sort(table(obs_df$cell_type_ontology_term_id), decreasing = TRUE)[1:10]
cat(paste("The top 10 cell types where tissue_ontology_term_id ==", LUNG_TISSUE))
print(top_10)
```

You can also define much more complex value filters. For example:

* combine terms with `and` and `or`
* use the `%in%` operator to query on multiple values

```{r}
# You can also do more complex queries, such as testing for inclusion in a list of values
obs_df <- as.data.frame(human$obs$read(
  column_names = c("cell_type_ontology_term_id"),
  value_filter = "tissue_ontology_term_id %in% c('UBERON:0002082', 'UBERON:OOO2084', 'UBERON:0002080')"
))

# Summarize
top_10 <- sort(table(obs_df$cell_type_ontology_term_id), decreasing = TRUE)[1:10]
print(top_10)
```

### Full census stats

This example queries all organisms in the Census, and summarizes the diversity of various metadata labels.

```{r}
cols_to_query <- c(
  "cell_type_ontology_term_id",
  "assay_ontology_term_id",
  "tissue_ontology_term_id"
)

total_cells <- 0
for (organism in census$get("census_data")$names()) {
  print(organism)
  obs_df <- as.data.frame(
    census$get("census_data")$get(organism)$obs$read(column_names = cols_to_query)
  )
  total_cells <- total_cells + nrow(obs_df)
  for (col in cols_to_query) {
    cat(paste("  Unique ", col, " values: ", length(unique(obs_df[[col]])), "\n", sep = ""))
  }
}
cat(paste("Complete cell census contains", total_cells, "cells."))
```
