---
title: "Access CELLxGENE collaboration embeddings (scVI, Geneformer)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Access CELLxGENE collaboration embeddings (scVI, Geneformer)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(ggplot2)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 88, max.print = 100)
```

This notebook demonstrates basic access to CELLxGENE collaboration embeddings of CELLxGENE Discover Census. Currently, embeddings from scVI and a fine-tuned Geneformer model are maintained by CELLxGENE Discover. There are other CELLxGENE-hosted embeddings contributed by the community to CELLxGENE Discover, find out more about these in the [Census model page](https://cellxgene.cziscience.com/census-models).

**Contents**

Querying cells and loading embeddings as:

1. Seurat reductions
2. Bioconductor SingleCellExperiment reductions
3. sparseMatrix

## Open Census

```{r,warning=FALSE,message=FALSE}
library("cellxgene.census")
census <- open_soma(census_version = "2023-12-15")
```

## Load embeddings as Seurat reductions

The high-level [`cellxgene.census::get_seurat()`](https://chanzuckerberg.github.io/cellxgene-census/r/reference/get_seurat.html) function can both query the Census and load embeddings into [dimensional reductions](https://satijalab.org/seurat/archive/v3.0/dim_reduction_vignette) of the Seurat object.

Here we will ask for a Seurat object with the expression data for all human cells of `tissue_general` equal to `'central nervous system'`, along with the scVI and geneformer embeddings (`obsm_layers`).

```{r,warning=FALSE,message=FALSE}
library("Seurat")

seurat_obj <- get_seurat(
  census, 
  organism = "homo_sapiens",
  obs_value_filter = "tissue_general == 'central nervous system'",
  obs_column_names = c("cell_type"),
  obsm_layers = c("scvi","geneformer")
)
```

With the embeddings stored as dimensional reductions on `seurat_obj`, we can take a quick look at the scVI embeddings in a 2D scatter plot via UMAP, colored by the Census `cell_type` annotations.

```{r,warning=FALSE,message=FALSE,fig.width=7,fig.height=4}
seurat_obj <- RunUMAP(
  seurat_obj, 
  reduction = "scvi",
  dims=1:ncol(Embeddings(seurat_obj, "scvi"))
)

DimPlot(seurat_obj, reduction = "umap", group.by = "cell_type") +
  theme(legend.text = element_text(size = 8))
```

## Load embeddings as SingleCellExperiment reductions

Similarly, [`cellxgene.census::get_single_cell_experiment()`](https://chanzuckerberg.github.io/cellxgene-census/r/reference/get_single_cell_experiment.html) can query the Census and store embeddings as dimensionality reduction results on a Bioconductor `SingleCellExperiment` object. 

```{r,warning=FALSE,message=FALSE}
library("SingleCellExperiment")
sce_obj <- get_single_cell_experiment(
  census, 
  organism = "homo_sapiens",
  obs_value_filter = "tissue_general == 'central nervous system'",
  obs_column_names = c("cell_type"),
  obsm_layers = c("scvi","geneformer")
)
```

Then, we can view a UMAP of the Geneformer embeddings colored by `cell_type`.

```{r,warning=FALSE,message=FALSE,fig.width=7,fig.height=4}
sce_obj <- scater::runUMAP(sce_obj, dimred = "geneformer")
scater::plotReducedDim(sce_obj, dimred = "UMAP", colour_by = "cell_type")
```

## Load embeddings as `sparseMatrix`

Lastly, we can use a [SOMAExperimentAxisQuery](https://single-cell-data.github.io/TileDB-SOMA/reference/SOMAExperimentAxisQuery.html#method-read-) for lower-level access to the embeddings' numerical data. This can be more performant for some use cases that don't need the other features of Seurat or SingleCellExperiment.

```{r,warning=FALSE,message=FALSE}
query <- census$get("census_data")$get("homo_sapiens")$axis_query(
  "RNA", obs_query = tiledbsoma::SOMAAxisQuery$new(value_filter = "tissue == 'tongue'")
)
embeddings <- query$to_sparse_matrix("obsm", "geneformer")
str(embeddings)
```

Each row of the embeddings `sparseMatrix` provides the fine-tuned Geneformer model's 512-dimensional embedding vector for a cell, with the cell `soma_joinid`s in the row names. With different arguments, [`SOMAExperimentAxisQuery$to_sparse_matrix()`](https://single-cell-data.github.io/TileDB-SOMA/reference/SOMAExperimentAxisQuery.html#method-SOMAExperimentAxisQuery-to_sparse_matrix) can also be read the scVI embeddings or the expression data.

Still lower-level access is available through [`SOMAExperimentAxisQuery$read()`](https://single-cell-data.github.io/TileDB-SOMA/reference/SOMAExperimentAxisQuery.html#method-read-), which streams Arrow tables. And other methods on `SOMAExperimentAxisQuery` can fetch metadata like `cell_type`:

```{r}
head(as.data.frame(query$obs(column_names = c("soma_joinid","cell_type"))$concat()))
```

The `SOMAExperimentAxisQuery` loads only what you ask for from the Census, unlike the high-level `get_seurat()` and `get_single_cell_experiment()` functions, which eagerly populate those objects based on your query.

```{r}
census$close()
```
